Effectful Semantics for Substructural SSA

### 2023-11-14 pldi24/paper.tex
Tap

To solve this problem, we need to introduce the concept of a
\textit{label-substitution} \(\mc{L}\), which we define as a function that takes
labels \(\lbl{\ell}\) and variables \(x\) to blocks \(t\) such that, for any
fresh variable \(x\), \([a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)\). We
introduce the judgement \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) to state that the
label-substitution \(\mc{L}\) takes the label-set \(\ms{L}\) to the label-set
\(\ms{K}\); that is, for each \(\lbl{\ell}[\Gamma](x: A) \in \ms{L}\), we have
that \(\mc{L}(\lbl{\ell}, x)\) is a block that branches to \(\ms{K}\). This is
formally defined via the rules \rle{lb-subst-nil}, which states that this is
vacuously true for \(\ms{L} = \cdot\), and \rle{lb-subst-cons}, which says that
if \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma,
\thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}\) then \(\lbsubst{\mc{L}}{\ms{L},
\lhyp{\lbl{\ell}}{\Gamma}{A}}{\ms{K}}\), in Figure \ref{fig:ssa-subst-typing}.
Given a context \(\gamma\) and a label-set \(\ms{L}\), we can convert it to a
label-substitution \(\substlbs{\gamma}{\ms{L}}\) using the definition in Figure
\ref{fig:ssa-subst-ops}: using this, we can state the full substitution theorem
as follows:
\begin{lemma}[Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then
    \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\) and
    \(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}\)
    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such
    that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and
    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
    \(\haslb{\Theta}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}\)
    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
    \ms{W}'\) such that 
    \begin{itemize}
      \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
      \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
      \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
    \end{itemize}
  \end{itemize}
\end{lemma}

@@

\TODO{label substitution text}

\TODO{some text about rewriting}

>>>>

To solve this problem, we need to introduce the concept of a
\textit{label-substitution} \(\mc{L}\), which we define as a function that takes
labels \(\lbl{\ell}\) and variables \(x\) to blocks \(t\) such that, for any
fresh variable \(x\), \([a/x]\mc{L}(\lbl{\ell}, x) = \mc{L}(\lbl{\ell}, a)\). We
introduce the judgement \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) to state that the
label-substitution \(\mc{L}\) takes the label-set \(\ms{L}\) to the label-set
\(\ms{K}\); that is, for each \(\lbl{\ell}[\Gamma](x: A) \in \ms{L}\), we have
that \(\mc{L}(\lbl{\ell}, x)\) is a block that branches to \(\ms{K}\). This is
formally defined via the rules \rle{lb-subst-nil}, which states that this is
vacuously true for \(\ms{L} = \cdot\), and \rle{lb-subst-cons}, which says that
if \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\) and \(\haslb{\Gamma,
\thyp{x}{A}{}}{\mc{L}(\lbl{\ell}, x)}{\ms{K}}\) then \(\lbsubst{\mc{L}}{\ms{L},
\lhyp{\lbl{\ell}}{\Gamma}{A}}{\ms{K}}\), in Figure \ref{fig:ssa-subst-typing}.
Given a context \(\gamma\) and a label-set \(\ms{L}\), we can convert it to a
label-substitution \(\substlbs{\gamma}{\ms{L}}\) using the definition in Figure
\ref{fig:ssa-subst-ops}: using this, we can state the full substitution lemma as
follows:
\begin{lemma}[Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item If \(\hasty{\Gamma}{p}{a}{A}{q}\), then
    \(\hasty{\Theta}{p}{[\gamma]a}{A}{q}\) and
    \(\hasty{\Theta}{p}{\exprletsubst{\substctx{\gamma}{\Gamma}}{a}}{A}{q}\)
    \item If \(\haslb{\Gamma}{t}{\ms{L}}\), then there exists \(\ms{K}\) such
    that \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) and
    \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
    \(\haslb{\Theta}{\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}}{\ms{K}}\)
    \item If \(\lhaslb{\ms{W}}{L}{\ms{L}}\), then there exists \(\ms{K},
    \ms{W}'\) such that 
    \begin{itemize}
      \item \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\) 
      \item \(\lbsubst{\substlbs{\gamma}{\ms{W}'}}{\ms{W}}{\ms{W}'}\) 
      \item \(\lhaslb{\ms{W}'}{[\gamma]L}{\ms{K}}\)
    \end{itemize}
  \end{itemize}
\end{lemma}

@@

We can also, using the rules in Figure \ref{fig:ssa-subst-ops}, define
capture-avoiding label-substitution for blocks \([\mc{L}]t\) and label-sets
\([\mc{L}]L\), allowing us to state the following lemma:
\begin{lemma}[Label Substitution] 
  Given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), 
  \begin{itemize}
    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\),
    \(\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}\)
    \item For all \(\lhaslb{\ms{W}}{L}{\ms{L}}\),
    \(\lhaslb{\ms{W}}{[\mc{L}]L}{\ms{K}}\)
  \end{itemize}
\end{lemma}

\TODO{some text about rewriting?}
